\pkg{cleanR}: Maid for Cleaning Datasets in \proglang{R}

### 2016-12-21 latex/article_vol2.tex
Fixed data

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{data.table}, and \pkg{plyr} have
made data wrangling a lot easier, but only a handful of packages such
as \pkg{editrules}, \pkg{validate}, \pkg{DataCombine}, and
\pkg{janitor} attempt to implement systematic, reproducible data
cleaning.  These packages use different approaches for data cleaning:
\pkg{editrules} and \pkg{validate} provide frameworks for setting up
and checking constraints on the variables, while \pkg{DataCombine} and
\pkg{janitor} both provide a few functions for identifying problems
(e.g, duplicates, dates coded as numbers, etc.) in data.

@@

\begin{figure}
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{cleanR_toyData.pdf}} 
\frame{\includegraphics[width=7.5cm,page=3]{cleanR_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{cleanR_testData.pdf}
\end{center}
\label{fig:example1}
\caption{Example output from running \R{clean()} on the \R{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\end{figure}

@@

Though the \R{clean()} function is very easy to use, it should not be
mistaken to be inflexible. A large number of function arguments allows
for the cleaning overview document to be molded according to the
user's needs. The most common arguments are summarized in Table~\ref{}
but in order to understand this distinction, a brief glimpse of the
inner structure of \R{clean()} might be useful. This can be obtained
from Figure~\ref{figure:cleanStructure}.

Below, we present a selection of these arguments in
sections \hl{xx} and \hl{xxx}. We do this under two distinct headlines
to emphasize that two levels of control are available when using
\R{clean()}. Either, we can control the \hl{[something] - essentially,
  this is everything but SVC-functions - } or we can control what
quality assessments and descriptions each variable are subjected
to.  \hl{segway to next bit or
  just something to end the paragraph}

\begin{figure}
\begin{description}
\item[Input] A dataset
	\begin{itemize} 
		\item This should be of type \R{data.frame} or \R{tibble} \hl{or data.table?}
	\end{itemize}
\item[Create contents] For each variable in the dataset, do the following:
	\begin{description}
		\item[Stage 1:] Pre-checks 
			\begin{itemize}
				\item Is the variable suitable for summarization, visualization and checks?
				\begin{description}
					\item[Yes:] Go to stage 2. 
					\item[No:] Move on to the next variable.
				\end{description}
			\end{itemize}
		\item[Stage 2:] SVC-steps
			\begin{description}
				\item[Summarize] Call \R{summarize()} to produce a summary table describing the variable. What features enter this table depends on the data class of the variable.
				\item[Visualize]  Call \R{visualize()} to produce a plot visualizing the distribution of the variable. 
				\item[Check]  Call \R{check()} to apply quality- and error checks to the variable. What checks are used depends on the data class of the variable.
			\end{description}
	\end{description}
\item[Output] Files for a overview document are saved to the disc and possibly also opened.
	\begin{itemize}
		\item Always a \R{rmarkdown} (.Rmd) file
		\item Possible also a html or pdf file
	\end{itemize}
\end{description}
\caption{\hl{quite odd having a figure like this? maybe do a flowchart?}}
\label{figure:cleanStructure}
\end{figure}

\subsection{Controlling [something]}
\label{subsection:controlSomething}

@@

\smallskip Control \hl{...eh?}\\
\quad \R{useVar} & What variables should be used? & \R{NULL} (corresponding to all variables) \\
\quad \R{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \R{"asIs"} \\
\quad \R{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \R{FALSE} \\
\quad \R{listChecks} & Should an overview of what checks were performed by listed in the \textit{Data cleaning summary}? &  \R{TRUE} \\

\smallskip Control SVC-step \\
\quad \R{mode} & What steps should be performed for each variable (out of the three possibilities \textit{summarize}, \textit{visualize}, \textit{check})? & \R{c("summarize", "visualize", "check")} \\
\quad \R{labelled\_as} & How should variables of class \R{labelled} be handled (as factors, is missing values or by ignoring labels)? & \R{"factor"} \\
\quad \R{smartNum} & Should numerical values with only a few unique levels be flagged an treated as factor variables? & \R{TRUE} \\
\quad \R{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \R{Inf} \\
\quad \R{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \R{2} \\
\quad \R{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \R{TRUE} \\

\smallskip Control output \\
\quad \R{output} & Type of output file to be produced (html, pdf or \hl{?? what does screen mean in the documentation??}) & \R{"pdf"} \\
\quad \R{render} & Should the output file be rendered from markdown? & \R{TRUE} \\
\quad \R{openResult} & If a  pdf/html file is rendered, should it automatically open afterwards, and if not, should the \R{rmarkdown} file? & \R{TRUE} \\

\hline
\end{tabular}
\caption{A selection of commonly used\hl{/useful?} arguments of \R{clean()}.  \hl{Note: I have everything here except \R{file} and \R{quiet}, replace, vol, standAlone + stuff related to selecting SVC-functions (dealt with in next subsection)}.  \hl{Fix formatting for \R{onlyProblematic}.}}
\label{table.cleanFormals}
\end{table}

\hl{Describe arguments in Table \ref{table.cleanFormals}. Maybe do a few examples of interesting argument combinations?}

\subsection{Something about what check, visual and summary functions are available}

@@

@@

\section{Something about the interactive cleanR mode}
While overview documents are great for documenting our work, they might not feel very natural for actually performing data cleaning or data wrangling based on the results we found. Therefore, \pkg{cleanR} also provides more standard \R{R} interactive tools, such as functions that print results to the console or saves it in a variable for later use. Or, more precisely, the functions used by \R{clean()} to do actual data summaries, visualizations and checks, as described above, can also be used interactively. \hl{segway}

\hl{Maybe write here explicitly that this section introduces the three functions: \R{summarize()}, \R{check()} and \R{visualize()}?}

\subsection{Cleaning by hand: An example}
Let's say we wish to look further into a certain variable from \R{toyData}, namely \R{var2}. The data cleaning summary found some issues in this variable, and we would like to recall what these issues were. This can be done using the command
\begin{Schunk} 
\begin{Sinput} 
> check(toyData$var2)
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
$identifyOutliers
Note that the following possible outlier values were detected: 82, 999. $
\end{Soutput}
\end{Schunk}
\hl{delete that last \$, it's only there to make my TeX editor stop highlighting everything. Note: \$s cannot be escaped here, as Schunk is verbatim.}

Note that the arguments specifying which checks to perform, as described in \hl{XXXX} above, are in fact passed to \R{check()}, and thus they can also be used here. For instance, if we only want the result of the check for miscoded missing values, we write
\begin{Schunk}
\begin{Sinput}
> check(toyData$var2, numericChecks = "identifyMissing")
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}
An equivalent way to call only a single, specific \R{checkFunction} is by using it directly on the variable, i.e. 
\begin{Schunk}
\begin{Sinput}
> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN. $
\end{Soutput}
\end{Schunk}
\hl{delete that last \$}

@@

\begin{Schunk}
\begin{Sinput}
> missVar2 <- identifyMissing(toyData$var2)
> str(missVar2)
\end{Sinput}
\begin{Soutput}
List of 3
 $ problem      : logi TRUE
 $ message      : chr "The following suspected missing value codes enter as 
 		 regular values: \\\"999\\\", \\\"NaN\\\"."
 $ problemValues: num [1:2] 999 NaN
 - attr(*, "class")= chr "checkResult"
\end{Soutput}
\end{Schunk}
The most important thing to note here is that while the printed message is made for easy reading, the actual values of the variable causing the issue are still obtainable. If we for instance decide that the values \R{999} and \R{NaN} in \R{var2} are in fact miscoded missing values, we can easily replace them with \R{NA}s:
\begin{Schunk}
\begin{Sinput}
> toyData$var2[toyData$var2 %in% missVar2$problemValues] <- NA
> identifyMissing(toyData$var2}
\end{Sinput}
\begin{Soutput}
No problems found. $
\end{Soutput}
\end{Schunk}
\hl{remove that last \$}. \hl{This is quite a practical example, but it essentially also gives the basic structure for misusing cleanR by generating a loop that does autocleaning... What do you think? Perhaps those who know how to make loops also know that they shouldn't do automatic data cleaning?}

@@

\begin{table}[ht]
\small
\begin{tabular}{p{0.2\linewidth}p{0.2\linewidth}p{0.2\linewidth}p{0.2\linewidth}}
& \R{summaryFunction} & \R{visualFunction} & \R{checkFunction} \\
\hline
Input (required) & \R{v} - a variable vector \newline \R{...}  &  \R{v} - a variable vector \newline \R{vnam} - the variable name (as character string) \newline \R{doEval} - a logical (\R{TRUE}/\R{FALSE}) controlling the output type of the function & \R{v} - a variable vector \newline \R{nMax} - an integer (or \R{Inf}), controlling how many problematic values are printed, if relevant \newline \R{...}  \\
Input (optional) &  \R{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \R{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g. a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries \newline \R{\$feature} - a label for the summary value (as character string) \newline \R{\$result} - the result of the summary (as character string) & A character string with \R{R} code for producing a plot. This code should be standalone, i.e. should include the data if necessary. & A list with entries \newline \R{\$problem} - a logical identifying whether an issue was found \newline \R{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \R{rmarkdown} \\
Output (recommended) & A \R{summaryResult} object (i.e. an attributed list with entries \R{\$feature}, \R{\$result} and \R{\$value}, the latter being the values from \R{\$result} in their original format). & \begin{tabular}{p{0.4\linewidth} | p{0.4\linewidth}} If \R{doEval} is \R{TRUE}: \newline A plot that will be opened by the graphic device in \R{R}. & If \R{doEval} is \R{FALSE}: \newline   A text string with \R{R} code, as described above. \end{tabular} & A \R{checkResult} object (an attributed list with entries \R{\$problem}, \R{\$message} and \R{\$problemValues}, the latter being either \R{NULL} or the problem causing values, as they were found in \R{v}, whichever is relevant. \\
Tools available for producing the function & \R{summaryResult()} & - & \R{messageGenerator()} \newline \R{checkResult()} 
\end{tabular}
\caption{\hl{blabla. Fix formatting! Maybe use multirow and raggedRight stuff?}}
\label{table.functionTypes}
\end{table}

>>>>

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr}, and \pkg{data.table} have
made data wrangling a lot easier, but only a handful of packages such
as \pkg{editrules}, \pkg{validate}, \pkg{DataCombine}, and
\pkg{janitor} attempt to implement systematic, reproducible data
cleaning.  These packages use different approaches for data cleaning:
\pkg{editrules} and \pkg{validate} provide frameworks for setting up
and checking constraints on the variables, while \pkg{DataCombine} and
\pkg{janitor} both provide a few functions for identifying problems
(e.g, duplicates, dates coded as numbers, etc.) in data.

@@

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{cleanR_toyData.pdf}} 
\frame{\includegraphics[width=7.5cm,page=3]{cleanR_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{cleanR_testData.pdf}
\end{center}
\label{fig:example1}
\caption{Example output from running \R{clean()} on the \R{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\end{figure}

@@

Though the \R{clean()} function is very easy to use, it should not be
mistaken to be inflexible. A large number of function arguments allows
for the cleaning overview document to be molded according to the
user's needs.

The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the cleaning process they influence. In order to understand
this distinction, a glimpse of the inner structure of \R{clean()} is
shown in Figure~\ref{figure:cleanStructure}. In the following we give
examples on how to use these parameters to influence the output of
\R{clean() }.

@@

\smallskip Control input variable and summary\\
\quad \R{useVar} & What variables should be used? & \R{NULL} (corresponding to all variables) \\
\quad \R{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \R{"asIs"} \\
\quad \R{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \R{FALSE} \\
\quad \R{listChecks} & Should an overview of what checks were performed by listed in the \textit{Data cleaning summary}? &  \R{TRUE} \\

\smallskip Control summarize, visualize, and check steps \\
\quad \R{mode} & What steps should be performed for each variable (out of the three possibilities \textit{summarize}, \textit{visualize}, \textit{check})? & \R{c("summarize", "visualize", "check")} \\
\quad \R{labelled\_as} & How should variables of class \R{labelled} be handled (as factors, is missing values or by ignoring labels)? & \R{"factor"} \\
\quad \R{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \R{TRUE} \\
\quad \R{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \R{Inf} \\
\quad \R{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \R{2} \\
\quad \R{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \R{TRUE} \\

\smallskip Control output and post-processing \\
\quad \R{output} & Type of output file to be produced (html, or pdf) & \R{"pdf"} \\
\quad \R{render} & Should the output file be rendered from markdown? & \R{TRUE} \\
\quad \R{openResult} & If a  pdf/html file is rendered, should it automatically open afterwards, and if not, should the \R{rmarkdown} file? & \R{TRUE} \\

\hline
\end{tabular}
\caption{A selection of commonly used arguments to \R{clean()} separated into the parts they control.}
\label{table.cleanFormals}
\end{table}

% Below, we present a selection of these arguments in
% sections \hl{xx} and \hl{xxx}. We do this under two distinct headlines
% to emphasize that two levels of control are available when using
% \R{clean()}. Either, we can control the \hl{[something] - essentially,
%   this is everything but SVC-functions - } or we can control what
% quality assessments and descriptions each variable are subjected
% to.  \hl{segway to next bit or
%   just something to end the paragraph}

\begin{figure}[tb]
\begin{description}
\item[Input] A dataset
	\begin{itemize} 
		\item This should be of type \R{data.frame} or \R{tibble} \hl{or data.table?}
	\end{itemize}
\item[Create contents] For each variable in the dataset, do the following:
	\begin{description}
		\item[Stage 1:] Pre-checks 
			\begin{itemize}
				\item Is the variable suitable for summarization, visualization and checks?
				\begin{description}
					\item[Yes:] Go to stage 2. 
					\item[No:] Move on to the next variable.
				\end{description}
			\end{itemize}
		\item[Stage 2:] SVC-steps
			\begin{description}
				\item[Summarize] Call \R{summarize()} to produce a summary table describing the variable. What features enter this table depends on the data class of the variable.
				\item[Visualize]  Call \R{visualize()} to produce a plot visualizing the distribution of the variable. 
				\item[Check]  Call \R{check()} to apply quality- and error checks to the variable. What checks are used depends on the data class of the variable.
			\end{description}
	\end{description}
\item[Output] Files for a overview document are saved to the disc and possibly also opened.
	\begin{itemize}
		\item Always a \R{rmarkdown} (.Rmd) file
		\item Possible also a html or pdf file
	\end{itemize}
\end{description}
\caption{\hl{quite odd having a figure like this? maybe do a flowchart?}}
\label{figure:cleanStructure}
\end{figure}

% \subsection{Controlling [something]}
% \label{subsection:controlSomething}

For example, to get a summary document that only contains the
variables with potential problems, and with a limit of maximum 10
printed potential errors, we can write (output not shown)

\begin{Schunk}
\begin{Sinput}
> clean(toyData, onlyProblematic=TRUE, maxProbVals=10)
\end{Sinput}
\end{Schunk}

The final rendering of the generated

By default, \R{clean()} runs all the summary,
visualization, and check functions that are
implemented. Table~\ref{table.SVCfunctions} lists these functions but
we can also use the \R{allSummaryFunctions()},
\R{allVisualFunctions()}, and \R{allCheckFunctions()} function in
\proglang{R} to obtain a list. For example the implemented summary
functions are:

%\subsection{Something about what check, visual and summary functions are available}

@@

\begin{Schunk}
\begin{Sinput}
> allSummaryFunctions()
\end{Sinput}
\begin{Soutput}
----------------------------------------------------------------------------
name           description                     classes                      
-------------- ------------------------------- -----------------------------
centralValue   Compute median or mode          character, Date,
factor,     
                                               integer, labelled,
                                               logical,  
                                               numeric

countMissing   Compute ratio of missing        character, Date,
factor,     
               observations                    integer, labelled,
               logical,  
                                               numeric

minMax         Find minimum and maximum        integer, numeric, Date       
               values

quartiles      Compute 1st and 3rd quartiles   integer, numeric

uniqueValues   Count number of unique values   character, Date,
factor,     
                                               integer, labelled,
                                               logical,  
                                               numeric

variableType   Data class of variable          character, Date,
factor,     
                                               integer, labelled,
                                               logical,  
                                               numeric                      
----------------------------------------------------------------------------
\end{Soutput}
\end{Schunk}

@@

\section{Using cleanR interactively}
While overview documents are great for presenting and documenting the
data cleaning checks, it may be natural to work more interactively
through the data cleaning process. \pkg{cleanR} also
provides more standard \proglang{R} interactive tools, such as
functions that print results to the console or returns the information
as an object for later use. This section describes how to use the
functions \R{check()}, \R{summarize()} and \R{visualize()} to work
interactively with \pkg{cleanR}.

\subsection{Data cleaning by hand: An example}
Let's say we wish to look further into a certain variable from
\R{toyData}, namely \R{var2}. The data cleaning summary found some
issues in this variable, and we would like to recall what these issues
were. This can be done using the \R{check()} command
\begin{Schunk} 
\begin{Sinput} 
> check(toyData$var2)
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
$identifyOutliers
Note that the following possible outlier values were detected: 82, 999. 
\end{Soutput}
\end{Schunk}
% $

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \R{check()}, and
thus they can also be used here. For instance, if we only want the
result of the check for miscoded missing values, we write
\begin{Schunk}
\begin{Sinput}
> check(toyData$var2, numericChecks = "identifyMissing")
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}
The \R{numericChecks} argument XXX.. ... not defined

An equivalent way to call only a single, specific \R{checkFunction}
such as \R{identifyMissing} (see \ref{table.SVCfunctions} for a list
of check functions) is by using it directly on the variable, i.e.
\begin{Schunk}
\begin{Sinput}
> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN. 
\end{Soutput}
\end{Schunk}
% $

@@

\begin{Schunk}
\begin{Sinput}
> missVar2 <- identifyMissing(toyData$var2)
> str(missVar2)
\end{Sinput}
\begin{Soutput}
List of 3
 $ problem      : logi TRUE
 $ message      : chr "The following suspected missing value codes enter as 
 		 regular values: \\\"999\\\", \\\"NaN\\\"."
 $ problemValues: num [1:2] 999 NaN
 - attr(*, "class")= chr "checkResult"
\end{Soutput}
\end{Schunk}
The most important thing to note here is that while the printed
message is made for easy reading, the actual values of the variable
causing the issue are still obtainable in the element
\R{problemValues}. If we for instance decide that the values \R{999}
and \R{NaN} in \R{var2} are in fact miscoded missing values, we can
easily replace them with \R{NA}s:
\begin{Schunk}
\begin{Sinput}
> toyData$var2[toyData$var2 %in% missVar2$problemValues] <- NA
> identifyMissing(toyData$var2}
\end{Sinput}
\begin{Soutput}
No problems found.
\end{Soutput}
\end{Schunk}
% $

@@

\begin{table}[ht]
\scriptsize
\begin{tabular}{p{0.2\linewidth}p{0.2\linewidth}p{0.2\linewidth}p{0.2\linewidth}}
& \R{summaryFunction} & \R{visualFunction} & \R{checkFunction} \\
\hline
Input (required) & \R{v} - a variable vector \newline \R{...}  &  \R{v} - a variable vector \newline \R{vnam} - the variable name (as character string) \newline \R{doEval} - a logical (\R{TRUE}/\R{FALSE}) controlling the output type of the function & \R{v} - a variable vector \newline \R{nMax} - an integer (or \R{Inf}), controlling how many problematic values are printed, if relevant \newline \R{...}  \\
Input (optional) &  \R{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \R{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g. a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries \newline \R{\$feature} - a label for the summary value (as character string) \newline \R{\$result} - the result of the summary (as character string) & A character string with \R{R} code for producing a plot. This code should be standalone, i.e. should include the data if necessary. & A list with entries \newline \R{\$problem} - a logical identifying whether an issue was found \newline \R{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \R{rmarkdown} \\
Output (recommended) & A \R{summaryResult} object (i.e. an attributed list with entries \R{\$feature}, \R{\$result} and \R{\$value}, the latter being the values from \R{\$result} in their original format). & \begin{tabular}{p{0.4\linewidth} | p{0.4\linewidth}} If \R{doEval} is \R{TRUE}: \newline A plot that will be opened by the graphic device in \R{R}. & If \R{doEval} is \R{FALSE}: \newline   A text string with \R{R} code, as described above. \end{tabular} & A \R{checkResult} object (an attributed list with entries \R{\$problem}, \R{\$message} and \R{\$problemValues}, the latter being either \R{NULL} or the problem causing values, as they were found in \R{v}, whichever is relevant. \\
Tools available for producing the function & \R{summaryResult()} & - & \R{messageGenerator()} \newline \R{checkResult()} 
\end{tabular}
\caption{\hl{blabla. Fix formatting! Maybe use multirow and raggedRight stuff?}}
\label{table.functionTypes}
\end{table}
